D3064R0: OOTA Execution on Real Implementations is Provably Vacuous

### 2023-12-15 ootavacuous.tex
December 14 self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{quote}
	\co{x = ++i / ++i;}
\end{quote}

The standard does not specify the order of evaluation, so the resulting
value might be zero ($\frac{1}{2}$ truncated) or two ($\frac{2}{1}$).
This indeterminacy is why many coding guidelines prohibit side effects
in expressions, but the above code nevertheless conforms to the standard.

\subsubsection{Semantic Dependencies Can Be a Choice}
\label{sec:Semantic Dependencies Can Be a Choice}

But suppose that the implementation can prove that at any time that
the above statement might execute, \co{y} is equal to \co{z}?
That would allow the implementation to act as if the source code
was instead as follows:

@@

\subsubsection{Complex Semantic Dependencies Are a Choice}
\label{sec:Complex Semantic Dependencies Are a Choice}

@@

Therefore, non-speculative loads are prohibited from loading values from
speculative stores.

@@

This optimization process may, roughly speaking, be divided into two
activities, analysis of the code and classification of dependencies.

@@

>>>>

\begin{quote}
	\co{y = ++i / ++i;} \\
	\co{x = y * z;}
\end{quote}

The standard does not specify the order of evaluation, so the resulting
value of \co{y} might be zero ($\frac{1}{2}$ truncated) or two
($\frac{2}{1}$).
In the former case, there is no semantic dependency from \co{z} to
\co{x}, but in the latter case there is.
Thus, semantic dependencies are not just a function of a particular
execution through the source code, but also of arbitrary choices made
by the C++ implementation.\footnote{
	Thanks to Peter Sewell for pointing out this possibility.}

Note well that C++ implementations can and do act as if portions of
the source code was duplicated in order to apply specialization
optimizations, in which case any nondeterminism might also be
duplicated.

Many projects' coding guidelines prohibit side effects in expressions,
their goal being to obtain portable code whose behavior does not depend
on arbitrary choices on the part of the implementation.
However, the above code nevertheless conforms to the standard.

\subsubsection{Semantic Dependencies And Optimization}
\label{sec:Semantic Dependencies And Optimization}

But suppose that the implementation can prove that at any time that
the above statement might execute, \co{y} is equal to \co{z}?
That would allow the implementation to optimize, that is, to act as if
the source code was instead as follows:

@@

\subsubsection{Complex Semantic Dependencies Analysis is a Choice}
\label{sec:Complex Semantic Dependencies Analysis is a Choice}

@@

Therefore, non-speculative loads are prohibited from loading values from
speculative stores.
And yes, this does mean that hardware can produce OOTA cycles during
speculative execution, but the hardware is required to prevent all such
speculative exection from ever committing.

@@

This optimization process may, roughly speaking, be divided into two
activities, analysis of the code and classification of dependencies.
Two sections on these topics are followed by sections on optimizations
and on consequences.

@@

\subsubsection{C++ Implementations: Optimization}
\label{sec:C++ Implementations: Optimization}

C++ implementations undertake a wide variety of optimizations, in
some cases taking advantage of nondeterminism and in others applying
the results of code analysis to produce faster code.
The changes can be profound, for example, with large computations being
replaced by simple constants.
Either way, the implementation is behaving as if the original program
had been written more optimally.

In such cases, one option is to carry out analysis of semantic dependencies
on the optimized code.
The seL4 project took this approach to the extreme of analyzing
the binary output~\cite{ThomasSewell2013L4binaryVerification}.

In C++ implementations using compilers, non-determinism of a given
(post-optimization) code path is ``frozen'' into the resulting binary,
which greatly simplifies analysis of semantic dependencies.
Other c++ implementations (for example, those using interpreters or JITs)
might need to directly model any nondeterminism.
This latter situation might motivate additional C++ projects to ban
nondeterminism in favor of portable code.
